---
title: 'Ordinations: Principal components analysis and correspondence analysis'
output:
  html_document:
    df_print: paged
---

## Pre-class work:

Load the environmental data associated with the arctic pollen data available from the palaeoSig r-package.

```{r}
library(palaeoSig)
data(arctic.env)
```

Run a principal components analysis on the environmental data related to temperature, precipitation and sunshine (columns 9 through 49). Compare the function *princomp (stats)*, *prcomp (stats)*  and *rda (vegan)* and explore the effects of options *cor (princomp)*, *center (prcomp)*  and *scale (rda)* making screeplots using function *screeplot (stats)*.



## Class work:

Let’s reproduce results of pre-class work computing eigenvalue decompositions on covariance and correlation matrices of environmental data.

Functions *cov* and *cor* compute covariances and correlations among environmental variables and *eigen* computes eigenvalue decompositions.

We should be able to reproduce screeplots plotting eigenvalues *(myobject$values)*.

**_Compute eigenvalue decompositions on covariance and correlation matrices and plot the eigenvalues_**.   

The sum of all eigenvalues equals the total variance of our dataset. 

**_Compute variance explained by each principal component_**.

In class we discussed principal components scores. Principal components scores are obtained from an rda object using function *scores (vegan)*.

**_Extract principal components from the rda object using the correlation matrix_**.

To reproduce PC-scores we need eigenvectors. Eigenvectors are computed during eigenvalue decomposition and obtained via *myobject$vector*. The first column of this matrix is the first eigenvector. We need the first eigenvector to obtain PC-scores of axis 1. An eigenvector assigns a coefficient (species score) to each variable in a dataset. For one site/sample we have to multiply the coefficient with the according (scaled) environmental variable. Thereafter we have to sum over all products. This is the site/sample score of this specific  site/sample.

**_Compute PC1-scores manually._** 

Hint either 

```{r,eval=FALSE}
apply(...,1, ...  sum(mysample*eigenvector)
as.matrix(mymatrix)%*%eigenvector  (matrix multiplication)
```

**_Compare the two sets of PC1-scores_**

These scores are not identical due to different scalings, but their correlation is r = 1.

As an aside, for scaled variables/correlation matrices, multiplying an eigenvector with the square-root of its eigenvalue gives the correlation between variables and the PC. The sum of squared correlations is the eigenvalue of the PC. 


The goal of this slightly tedious exercise is to show that PCA is based on correlations/covariances among variables, some magic (eigenvalue decompositon) and then a simple linear combination of variables and some coefficients. 


## Correspondence analysis
 
Correspondence analysis is an ordination analysis closely related to PCA. The main difference is that PCA works with euclidean distances among samples and CA with chi-square distances. Another important difference is that PCA assumes a linear increase/decrease of a species along a gradient whereas CA allows for unimodal responses (i.e. increase and subsequent decrease). 
Mathematically, in CA the data matrix is pre-transformed and species and site scores are weighted. 

We will first have a look at a very nice dataset and subsequently deal with somewhat problematic datasets.

 
**Perfect dataset:**

TBA



**Problematic datset1: Abernethy forest**

```{r}
library(analogue)
data(abernethy)
abernethy2 <- abernethy[,!colnames(abernethy)%in%c('Age','Depth')]
``` 

**_Run pca on the Abernethy dataset, make a ordination biplot and describe this biplot._**
```{r,eval=FALSE}
rda()
plot(,scaling =3)
``` 


**_Run pca on the square root-transformed Abernethy dataset, make a ordination biplot and describe this biplot._**


**_Run ca on the square root-transformed Abernethy dataset, make a ordination biplot and describe this biplot._**
```{r,eval=FALSE}
cca()
plot(,scaling =3)
```  

Problematic datset2: Arctic pollen data

```{r}
data(arctic.pollen)
```

**_Run ca on the square root-transformed Arctic pollen dataset, make a ordination biplot and describe this biplot._**

For this dataset we can explain the … effect modelling temperature of the warmest months as function PC1 and PC2 scores using a (generalised) additive model.

**_Fit an ordination surface to the Arctic pollen ordination biplot._**

```{r,eval=FALSE}
ordisurf(myca,mydata$mtwa,add=TRUE)
```






